Capstone Movielens Report

### 2025-05-14 reports/capstone-movielens.site/02-0.methods.Rmd

```{r setup, include=FALSE, echo = FALSE}
r.path <- "../r"
src.folder <- "src"
support_scripts.folder <- "support-scripts"
support_functions.folder <- "support-functions"

r.src.path <- file.path(r.path, src.folder)
support_scripts.path <- file.path(r.src.path, support_scripts.folder)
support_functions.path <- file.path(r.src.path, support_functions.folder)
# Setup -----------------------------------------------------------------------
setup_script.file_path <- file.path(support_scripts.path,
                                               "setup.R")

# main_script.file_path <- file.path(r.src.path,
#                                    "capstone-movielens.main.R")
# stop(main_script.file_path)
stopifnot(file.exists(setup_script.file_path))

source(setup_script.file_path, 
       local = knitr::knit_global())
# or sys.source("your-script.R", envir = knitr::knit_global())

# External Common Helper functions -------------------------------------
# common_helper_functions.file_path <- file.path(support_functions.path,
#                                             "common-helper.functions.R")
# source(common_helper_functions.file_path, 
#        local = knitr::knit_global())

# knitr::knit_global()
#knitr::opts_chunk$set(echo = FALSE)
```

---
title: "Capstone Movielens Report"
author: "Azamat Kurbanaev"
date: "`r Sys.Date()`"
bibliography: references.bib
#csl: bit-numerical-mathematics.csl 
envir:
  parent.frame()
output: 
  bookdown::pdf_document2: 
    df_print: paged
    toc: true
    fig_caption: true
    keep_tex: true
    citation_package: biblatex
colorlinks: true
linkcolor: red
citecolor: blue
header-includes:
    - \usepackage{awesomebox}
    - \usepackage{hyperref}
    - \usepackage[natbib=true, 
                  style=numeric, 
                  backref=true, 
                  sorting=none]{biblatex}
    - \hypersetup{backref,
          pdfpagemode=Normal,
          colorlinks=true,
          implicit=false}
vignette: >
  %\VignetteIndexEntry{RMarkdown Citations - Numeric Style}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::knitr}
---

\newenvironment{infobox}[1]
  {
  \begin{itemize}
  \renewcommand{\labelitemi}{
    \raisebox{-.7\height}[0pt][0pt]{
      {\setkeys{Gin}{width=3em,keepaspectratio}
        \includegraphics{images/#1}}
    }
  }
  \setlength{\fboxsep}{1em}
  \begin{blackbox}
  \item
  }
  {
  \end{blackbox}
  \end{itemize}
  }

# Introduction / Overview / Executive Summary

The goal of the project is to build a Recommendation System using a [10M version of the MovieLens dataset](http://grouplens.org/datasets/movielens/10m/).
Following the [Netflix Grand Prize Contest](https://archive.nytimes.com/bits.blogs.nytimes.com/2009/09/21/netflix-awards-1-million-prize-and-starts-a-new-contest/index.html) requirements, we will evaluate the _Root Mean Square Error_ (_RMSE_) score, which, as shown in [Section 23.2 Loss function](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#sec-netflix-loss-function) of the _Course Textbook_, is defined as:
$$
\mbox{RMSE} = \sqrt{\frac{1}{N} \sum_{i,j}^{N} (y_{i,j} - \hat{y}_{i,j})^2}
$$

with $N$ being the number of user/movie combinations for which we make predictions and the sum occurring over all these combinations[@IDS2_23-2].

Our goal is to achieve a value of less than 0.86490 (compare with the _Netflix Grand Prize_ requirement: of at least 0.8563[@BigChaosSln]). 
\newpage

## Datasets Overview

To start with we have to generate two datasets derived from the _MovieLens_ one mentioned above:

* **edx:** we use it to develop and train our algorithms;
* **final_holdout_test:**  according to the course requirements, we use it exclusively to evaluate the _**RMSE**_ of our final algorithm.

For this purpose the following package has been developed by the author of this report: `edx.capstone.movielens.data`. The source code of the package is available [on GitHub](https://github.com/AzKurban-edX-DS/edx.capstone.movielens.data)[@edx_capstone_movielens_data].

Let's install the development version of this package from the GitHub repository and attach the correspondent library to the global environment:
```{r eval=FALSE }
if(!require(edx.capstone.movielens.data)) pak::pak("AzKurban-edX-DS/edx.capstone.movielens.data")

library(edx.capstone.movielens.data)
edx <- edx.capstone.movielens.data::edx
final_holdout_test <- edx.capstone.movielens.data::final_holdout_test
```

Now, we have the datasets listed above:
```{r }
summary(edx)
```

```{r }
summary(final_holdout_test)
```
\newpage

### `edx` Dataset
\
Let's look into the details of the `edx` dataset:
``` {r echo = TRUE}
str(edx)
```
Note that we have 9000055 rows and six columns in there:  
```{r }
dim_edx <- dim(edx)
print(dim_edx)
```

First, let's note that we have 10677 different movies: 
```{r}
n_movies <- n_distinct(edx$movieId)
print(n_movies)
```
and 69878 different users in the dataset:
```{r}
n_users <- n_distinct(edx$userId)
print(n_users)
```

Now, note the expressions below which confirm the fact explained in [Section _23.1.1 Movielens data_](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#movielens-data) of the _Course Textbook_[@IDS2] that not every user rated every movie:
```{r}
max_possible_ratings <- n_movies*n_users
sprintf("Maximum possible ratings: %s", max_possible_ratings)
sprintf("Rows in `edx` dataset: %s", dim_edx[1])
sprintf("Not every movie was rated: %s", max_possible_ratings > dim_edx[1])

As also explained in that section, we can think of these data as a very large matrix, with users on the rows and movies on the columns, with many empty cells. Therefore, we can think of a recommendation system as filling in the `NA`s in the dataset for the movies that some or all the users do not rate. A sample from the `edx` data below illustrates this idea[@IDS2_23-1-1]: 
```{r}
keep <- edx |> 
  dplyr::count(movieId) |> 
  top_n(4, n) |> 
  pull(movieId)

tab <- edx |> 
  filter(movieId %in% keep) |> 
  filter(userId %in% c(13:20)) |> 
  select(userId, title, rating) |> 
  mutate(title = str_remove(title, ", The"),
         title = str_remove(title, ":.*")) |>
  pivot_wider(names_from = "title", values_from = "rating")

print(tab)
```

The following plot of the matrix for a random sample of 100 movies and 100 users with yellow indicating a user/movie combination for which we have a rating shows how _sparse_ the matrix is:
```{r sparsity-of-movie-recs, echo=TRUE, fig.width=3, fig.height=3, out.width="40%"}
users <- sample(unique(edx$userId), 100)

rafalib::mypar()
edx|> 
  filter(userId %in% users) |> 
  select(userId, movieId, rating) |>
  mutate(rating = 1) |>
  pivot_wider(names_from = movieId, values_from = rating) |> 
  (\(mat) mat[, sample(ncol(mat), 100)])() |>
  as.matrix() |> 
  t() |>
  image(1:100, 1:100, z = _ , xlab = "Movies", ylab = "Users")
```

Further observations highlighted there that, as we can see from the distributions the author presented, some movies get rated more than others, and some users are more active than others in rating movies:
```{r movie-id-and-user-hists, echo=TRUE, fig.width=6, fig.height=3}
p1 <- edx |> 
  count(movieId) |> 
  ggplot(aes(n)) + 
  geom_histogram(bins = 30, color = "black") + 
  scale_x_log10() + 
  ggtitle("Movies")

p2 <- edx |> 
  count(userId) |> 
  ggplot(aes(n)) + 
  geom_histogram(bins = 30, color = "black") + 
  scale_x_log10() + 
  ggtitle("Users")

gridExtra::grid.arrange(p2, p1, ncol = 2)
```

Finally, we can see that no movies have a rating of 0. Movies are rated from 0.5 to 5.0 in 0.5 increments:
```{r }
#library(dplyr)
s <- edx |> group_by(rating) |>
  summarise(n = n())
print(s)
```

Further analysis of the `edx` dataset have been also inspired by the article mentioned above[@MRS-R-BEST], from which the code and explanatory notes below were cited.
\newpage

#### Rating distribution plot[@MRS-R-BEST]
\
The code below demonstrates another way of visualizing the rating distribution:
```{r}
edx |>
  group_by(rating) |>
  summarize(count = n()) |>
  ggplot(aes(x = rating, y = count)) +
  geom_bar(stat = "identity", fill = "#8888ff") +
  ggtitle("Rating Distribution") +
  xlab("Rating") +
  ylab("Occurrences Count") +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(n.breaks = 10) +
  theme_economist() +
  theme(axis.title.x = element_text(vjust = -5, face = "bold"), 
        axis.title.y = element_text(vjust = 10, face = "bold"), 
        plot.margin = margin(0.7, 0.5, 1, 1.2, "cm"))

This graph is another confirmation of what we found out above: rounded ratings occur more often than half-stared ones. The upward trend previously discussed is now perfectly clear, although it seems to top right between the 3 and 4-star ratings lowering the occurrences count afterward. That might be due to users being more hesitant to rate with the highest mark for whichever reasons they might hold[@MRS-R-BEST].
\newpage

#### Ratings per movie
\
##### Movie popularity count[@MRS-R-BEST]
\
```{r}
print(edx |> 
  group_by(movieId) |> 
  summarize(count = n()) |>
  slice_head(n = 10)
)
```

```{r}
summary(edx |> group_by(movieId) |> summarize(count = n()) |> select(count))
```
\newpage

##### Ratings per movie plot[@MRS-R-BEST]
\
```{r}
edx |>
  group_by(movieId) |>
  summarize(count = n()) |>
  ggplot(aes(x = movieId, y = count)) +
  geom_point(alpha = 0.2, color = "#4020dd") +
  geom_smooth(color = "red") +
  ggtitle("Ratings per movie") +
  xlab("Movies") +
  ylab("Number of ratings") +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(n.breaks = 10) +
  theme_economist() +
  theme(axis.title.x = element_text(vjust = -5, face = "bold"), 
        axis.title.y = element_text(vjust = 10, face = "bold"), 
        plot.margin = margin(0.7, 0.5, 1, 1.2, "cm"))
```

##### Movies' rating histogram[@MRS-R-BEST]
\
```{r}
edx |>
  group_by(movieId) |>
  summarize(count = n()) |>
  ggplot(aes(x = count)) +
  geom_histogram(fill = "#8888ff", color = "#4020dd") +
  ggtitle("Movies' rating histogram") +
  xlab("Rating count") +
  ylab("Number of movies") +
  scale_y_continuous(labels = comma) +
  scale_x_log10(n.breaks = 10) +
  theme_economist() +
  theme(axis.title.x = element_text(vjust = -5, face = "bold"), 
        axis.title.y = element_text(vjust = 10, face = "bold"), 
        plot.margin = margin(0.7, 0.5, 1, 1.2, "cm"))
```
\newpage

#### Ratings per user[@MRS-R-BEST]
\
##### User rating count (activity measure)
\
```{r}
print(edx |> 
  group_by(userId) |> 
  summarize(count = n()) |>
  slice_head(n = 10)
)
```

##### User rating summary
\
```{r}
summary(edx |> group_by(userId) |> summarize(count = n()) |> select(count))
```

##### Ratings per user plot
\
```{r}
edx |>
  group_by(userId) |>
  summarize(count = n()) |>
  ggplot(aes(x = userId, y = count)) +
  geom_point(alpha = 0.2, color = "#4020dd") +
  geom_smooth(color = "red") +
  ggtitle("Ratings per user") +
  xlab("Users") +
  ylab("Number of ratings") +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(n.breaks = 10) +
  theme_economist() +
  theme(axis.title.x = element_text(vjust = -5, face = "bold"), 
        axis.title.y = element_text(vjust = 10, face = "bold"), 
        plot.margin = margin(0.7, 0.5, 1, 1.2, "cm"))
```
\newpage

##### Users' rating histogram
\
```{r}
edx |>
  group_by(userId) |>
  summarize(count = n()) |>
  ggplot(aes(x = count)) +
  geom_histogram(fill = "#8888ff", color = "#4020dd") +
  ggtitle("Users' rating histogram") +
  xlab("Rating count") +
  ylab("Number of users") +
  scale_y_continuous(labels = comma) +
  scale_x_log10(n.breaks = 10) +
  theme_economist() +
  theme(axis.title.x = element_text(vjust = -5, face = "bold"), 
        axis.title.y = element_text(vjust = 10, face = "bold"), 
        plot.margin = margin(0.7, 0.5, 1, 1.2, "cm"))
```
\newpage

@@

::: {.noteblock data-latex=""}
The full source code of these functions is available in the [Logging Helper functions](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L20) section of the [capstone-movielens.main.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R) script on _GitHub_.
:::
\newpage

@@

We will split the `edx` dataset into a training set, which we will use to build and train our models, and a test set in which we will compute the accuracy of our predictions, the way described in [Section 23.1.1 Movielens data](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#movielens-data) of the _Course Textbook_ mentioned above[@IDS2_23-1-1].We will also use the _5-Fold Cross Validation_ method as described in [Section 29.6 Cross validation](https://rafalab.dfci.harvard.edu/dsbook-part-2/ml/resampling-methods.html#cross-validation) of the _Course Textbook_. 
To prepare datasets for processing, we will use the following functions, specifically designed for these operations:
```{r eval=FALSE }
make_source_datasets()
init_source_datasets()
```

@@

### The `make_source_datasets` function
\

@@

::: {.noteblock data-latex=""}
This code snippet is a [part](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R#L140) of the [make_source_datasets](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R#L87) _function_ code described above.
:::

Note that we used the `sample_train_validation_sets` function call to split the original dataset (`edx` in this case): 
```{r eval=FALSE}
    split_sets <- edx |>
      sample_train_validation_sets(fold_i*1000)
```

@@

## Overall Mean Rating (Naive) Model
\

Let's begin our analysis by evaluating the simplest model described in [Section _23.3 The First Model_ of the *Course Textbook*](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#a-first-model), and then gradually refine it through further research.
It is about a model that assumes the same rating for all movies and users with all the differences explained by random variation would look as follows:

$$
Y_{i,j} = \mu + \varepsilon_{i,j}
$$

with $\varepsilon_{i,j}$ independent errors sampled from the same distribution centered at 0 and $\mu$ the _true_ rating for all movies.

We know that the estimate that minimizes the RMSE is the least squares estimate of $\mu$ and, in this case, is the average of all ratings:

```{r pressure, echo=TRUE}
mu <- mean(edx$rating)
print(mu)
```

If we predict all unknown ratings with $\hat{\mu}$, we obtain the following RMSE: 
```{r}
mu.MSEs <- naive_model_MSEs(mu)
data.frame(fold_No = 1:5, MSE = mu.MSEs) |>
  data.plot(title = "MSE resuls of the 5-fold CV method applied to the Overall Mean Rating Model",
              xname = "fold_No", 
              yname = "MSE")

mu.RMSE <- sqrt(mean(mu.MSEs))
mu.RMSE
```
::: {.noteblock data-latex=""}
For the _Mean Squared Error_ data visualization we used [data.plot](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R#L592) function] defined in the [Data Visualization](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R#L447) section of the [data.helper.function.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R) script.
:::

```{r eval=FALSE}
data.plot <- function(data, 
                      title, 
                      xname, 
                      yname, 
                      xlabel = NULL, 
                      ylabel = NULL,
                      line_col = "blue",
                      # scale = 1,
                      normalize = FALSE) {
  y <- data[, yname]

  if (normalize) {
    y <- y - min(y)
  }

  if (is.null(xlabel)) {
    xlabel = xname
  }
  if (is.null(ylabel)) {
    ylabel = yname
  }

  aes_mapping <- aes(x = data[, xname], y = y)

  data |> 
    ggplot(mapping = aes_mapping) +
    ggtitle(title) +
    xlab(xlabel) +
    ylab(ylabel) +
    geom_point() + 
    geom_line(color=line_col)
}
```

Here we also used [naive_model_MSEs](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/common-helper.functions.R#L54) function defined in the [common-helper.functions.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/common-helper.functions.R) script (already mentioned above) to compute _Mean Squared Errors_ using _5-Fold Cross Validation_ method:

```{r eval=FALSE }
naive_model_MSEs <- function(val) {
  sapply(edx_CV, function(cv_item){
    mse(cv_item$validation_set$rating - val)
  })
}
```

One more function, defined in the [same script](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/common-helper.functions.R), that we will need for further analysis of the current model, is the [naive_model_RMSE](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/common-helper.functions.R#L59) one:
```{r eval=FALSE }
naive_model_RMSE <- function(val){
  sqrt(mean(naive_model_MSEs(val)))
}
```
\newpage

### Ensure that `mu.RMSE` value is the best for the current model
\

If we plug in any other number, we will get a higher RMSE. Let's prove that by the following small investigation:
```{r }
  deviation <- seq(0, 6, 0.1) - 3

  deviation.RMSE <- sapply(deviation, function(delta){
    naive_model_RMSE(mu + delta)
  })
```

Let's make a quick investigation of the `deviation.RMSE` result we have just got: 
```{r}
data.frame(delta = deviation, 
           delta.RMSE = deviation.RMSE) |> 
data.plot(title = TeX(r'[RMSE as a function of deviation ($\delta$) from the Overall Mean Rating ($\hat{mu}$)]'),
              xname = "delta", 
              yname = "delta.RMSE", 
              xlabel = TeX(r'[$\delta$]'), 
              ylabel = "RMSE")

which_min_deviation <- deviation[which.min(deviation.RMSE)]
min_rmse = min(deviation.RMSE)

print_log1("Minimum RMSE is achieved when the deviation from the mean is: %1",
         which_min_deviation)

print_log1("Is the previously computed RMSE the best for the current model: %1",
         mu.RMSE == min_rmse)
```
```{r eval=FALSE}
RMSEs.ResultTibble.OMR <- RMSEs.ResultTibble |> 
  RMSEs.AddRow("Overall Mean Rating Model", mu.RMSE)
```
```{r}
RMSE_kable(RMSEs.ResultTibble.OMR)
```

To win the grand prize of $1,000,000, a participating team had to get an RMSE of at least 0.8563[@BigChaosSln]. So we can definitely do better![@IDS2_23-3]
\newpage

## User Effect Model
\

To improve our model let's now take into consideration user effects as explained in [Section _23.4 User effects_](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#user-effects) of the *Course Textbook*. 
If we visualize the average rating for each user the way the [the author](https://x.com/rafalab) shows, we can see that there is substantial variability in the average ratings across users: 
```{r}
hist(edx.user_mean_ratings$mean_rating, nclass = 30)
```

Following the author's further explanation, to account for this variability, we will use a linear model with a _treatment effect_  $\alpha_i$ for each user. The sum $\mu+\alpha_i$ can be interpreted as the typical rating user $i$ gives to movies. So we write the model as follows:

$$
Y_{i,j} = \mu + \alpha_i + \varepsilon_{i,j}
$$

Statistics textbooks refer to the $\alpha$s as treatment effects. In the Netflix challenge papers, they refer to them as _bias_[@IDS2_23-4; @MFT_RS].

As it is stated here[@IDS2_23-4], it can be shown that the least squares estimate $\hat{\alpha}_i$ is just the average of $y_{i,j} - \hat{\mu}$ for each user $i$. So we can compute them this way:
```{r eval=FALSE}
a <- rowMeans(y - mu, na.rm = TRUE)
```

These considerations alows us to compute a _User Mean Ratings_ the following way:
```{r eval=FALSE}
put_log("Computing Average Ratings per User (User Mean Ratings)...")
user.mean_ratings <- rowMeans(edx.mx, na.rm = TRUE)
user_ratings.n <- rowSums(!is.na(edx.mx))

edx.user_mean_ratings <- 
  data.frame(userId = names(user.mean_ratings), 
             mean_rating = user.mean_ratings,
             n = user_ratings.n)

put_log("User Mean Ratings have been computed.")
```
```{r}
str(edx.user_mean_ratings)
```

And then we compute a _User Effect_ this way:
```{r eval=FALSE}
put_log("Computing User Effect per users ...")
edx.user_effect <- edx.user_mean_ratings |>
  mutate(userId = as.integer(userId),
         a = mean_rating - mu)

put_log("A User Effect Model has been builded")
```
```{r}
par(cex = 0.7)
hist(edx.user_effect$a, 30, xlab = TeX(r'[$\hat{alpha}_{i}$]'),
     main = TeX(r'[Histogram of $\hat{alpha}_{i}$]'))

str(edx.user_effect)
```

::: {.noteblock data-latex=""}
The full source code of the _User Effect_ computation is available in the [Model building: User Effect](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L531) section of the [capstone-movielens.main.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R) script on _GitHub_.
:::

Finally, we are ready to compute the `RMSE` (additionally using the [clamp](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/common-helper.functions.R#L4) helper function we defined above to keep predictions in the proper range):
```{r eval=FALSE}
put_log("Computing the RMSE taking into account user effects...")
start <- put_start_date()
edx.user_effect.MSEs <- sapply(edx_CV, function(cv_fold_dat){
  cv_fold_dat$validation_set |>
    left_join(edx.user_effect, by = "userId") |>
    mutate(resid = rating - clamp(mu + a)) |> 
    pull(resid) |> mse()
})
put_end_date(start)

edx.user_effect.RMSE <- sqrt(mean(edx.user_effect.MSEs))

RMSEs.ResultTibble.UE <- RMSEs.ResultTibble.OMR |> 
  RMSEs.AddRow("User Effect Model", edx.user_effect.RMSE)
```
```{r}
data.frame(fold_No = 1:5, MSE = edx.user_effect.MSEs) |>
  data.plot(title = "MSE resuls of the 5-fold CV method applied to the User Effect Model",
            xname = "fold_No", 
            yname = "MSE")

RMSE_kable(RMSEs.ResultTibble.UE)
```

::: {.noteblock data-latex=""}
The full source code of the _User Effect Model RMSE_ computation is available in the [Compute RMSE for User Effect Model](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L655) section of the [capstone-movielens.main.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R) script on _GitHub_.
:::

## User+Movie Effect (UME) Model
\

In [23.5 Movie effects](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#movie-effects) section of the *Course Textbook* the author draws our attention to the fact that some movies are generally rated higher than others. He also explains that a linear model with a _treatment effect_ $\beta_j$ for each movie can be used in this case, which can be interpreted as movie effect or the difference between the average ranking for movie $j$ and the overall average $\mu$:

$$
Y_{i,j} = \mu + \alpha_i + \beta_j +\varepsilon_{i,j}
$$
The author then shows how to use an approximation by first computing the least square estimate $\hat{\mu}$ and $\hat{\alpha}_i$, and then estimating $\hat{\beta}_j$ as the average of the residuals $y_{i,j} - \hat{\mu} - \hat{\alpha}_i$:

```{r eval=FALSE}
b <- colMeans(y - mu - a, na.rm = TRUE)
```

Inspired by this idea, a few support functions were developed by the author of this report, which we will use for our further analysis.

### UME Model: Support Functions
\

::: {.noteblock data-latex=""}
The full source code of the functions described in this section is available in the [User+Movie Effect Model Functions](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R#L1) section of the [UM-effect.functions.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R) script on _GitHub_.
:::

#### `train_user_movie_effect` Function
\

We use this function to build and train our model using the `train_set` dataset:
```{r eval=FALSE}
train_user_movie_effect <- function(train_set, lambda = 0){
  if (is.na(lambda)) {
    stop("Function: train_user_movie_effect
`lambda` is `NA`")
  }

  UM.effect <- train_set |>
    left_join(edx.user_effect, by = "userId") |>
    mutate(resid = rating - (mu + a)) |> 
    group_by(movieId) |>
    summarise(b = mean_reg(resid, lambda), n = n())

  stopifnot(!is.na(mean(UM.effect$b)))
  UM.effect
}
```

::: {.noteblock data-latex=""}
The function described above accepts the `lambda` parameter, which we will need later for the _Regularization_ method. We also use the [mean_reg](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/common-helper.functions.R#L63) function call, which we will also need for the *Regularization*.
We will explain that later in the [Regularization Method] section. 
For now, we omit the `lambda` parameter, accepting its default value `lambda = 0`. In this case, the `mean_reg` function is equivalent to the simple `mean` one.
:::

```{r eval=FALSE}
# Regularization --------------------------------------------------------------
mean_reg <- function(vals, lambda = 0, na.rm = TRUE){
  if (is.na(lambda)) {
    stop("Function: mean_reg
`lambda` is `NA`")
  }

  names(lambda) <- NULL
  sums <- sum(vals, na.rm = na.rm)
  N <- ifelse(na.rm, sum(!is.na(vals)), length(vals))
  sums/(N + lambda)
}
```

#### `train_user_movie_effect.cv` Function
\

We use the [train_user_movie_effect.cv](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R#L17) function to build and train our model using the `5-Fold Cross Validation` method. Below, we provide the most important part of the code of that function:
```{r eval=FALSE}
train_user_movie_effect.cv <- function(lambda = 0){
# ...
  start <- put_start_date()
  user_movie_effects_ls <- lapply(edx_CV, function(cv_fold_dat){
    cv_fold_dat$train_set |> train_user_movie_effect(lambda)
  })
  put_end_date(start)
  put_log("Function: train_user_movie_effect.cv:
User+Movie Effect list have been computed")

  user_movie_effects_united <- union_cv_results(user_movie_effects_ls)

  user_movie_effect <- user_movie_effects_united |>
    group_by(movieId) |>
    summarise(b = mean(b), n = mean(n))
  # ...
  user_movie_effect
}

::: {.noteblock data-latex=""}
Here we use the function call [union_cv_results](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R#L432), which is defined in the script [common-helper.functions.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R), to aggregate the `Cross Validation` results.
:::
```{r eval=FALSE}
union_cv_results <- function(data_list) {
  out_dat <- data_list[[1]]

  for (i in 2:CVFolds_N){
    out_dat <- union(out_dat, 
                     data_list[[i]])
  }

  out_dat
}
```

#### `calc_user_movie_effect_MSE` Function
\

The code of the function [calc_user_movie_effect_MSE](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R#L55) defined in the [UM-effect.functions.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R) script to calculate the `Mean Squared Error (MSE)` of the `UME Model` for the given `Test Set` is provided below:
```{r eval=FALSE}
calc_user_movie_effect_MSE <- function(test_set, um_effect){
  mse.result <- test_set |>
    left_join(edx.user_effect, by = "userId") |>
    left_join(um_effect, by = "movieId") |>
    mutate(resid = rating - clamp(mu + a + b)) |> 
    pull(resid) |> mse()

  stopifnot(!is.na(mse.result))
  mse.result
}
```

#### `calc_user_movie_effect_MSE.cv` Function
\

The code of the function [calc_user_movie_effect_MSE.cv](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R#L72) defined in the [UM-effect.functions.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R) script to calculate the `5-Fold Cross Validation` MSE result of the `UME Model` is provided below:
```{r eval=FALSE}
calc_user_movie_effect_MSE.cv <- function(um_effect){
  put_log("Function: user_movie_effects_MSE.cv:
Computing the RMSE taking into account User+Movie Effects...")
  start <- put_start_date()
  user_movie_effects_MSEs <- sapply(edx_CV, function(cv_fold_dat){
    cv_fold_dat$validation_set |> calc_user_movie_effect_MSE(um_effect)
  })
  put_end_date(start)

  put_log1("Function: user_movie_effects_MSE.cv:
MSE values have been plotted for the %1-Fold Cross Validation samples.", 
           CVFolds_N)

  mean(user_movie_effects_MSEs)
}

#### `calc_user_movie_effect_RMSE` Function
\

The code of the function [calc_user_movie_effect_RMSE](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R#L51) defined in the [UM-effect.functions.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R) script to calculate the `Root Mean Squared Error (RMSE)` of the `UME Model` for the given `Test Set` is provided below:

```{r eval=FALSE}
calc_user_movie_effect_RMSE <- function(test_set, um_effect){
  mse <- test_set |> calc_user_movie_effect_MSE(um_effect)
  sqrt(mse)
}
```

#### `calc_user_movie_effect_RMSE.cv` Function
\

The code of the function [calc_user_movie_effect_RMSE.cv](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R#L65) defined in the [UM-effect.functions.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R) script to calculate the `5-Fold Cross Validation` RMSE result of the `UME Model` is provided below:

```{r eval=FALSE}
calc_user_movie_effect_RMSE.cv <- function(um_effect){
  user_movie_effects_MSE <- calc_user_movie_effect_MSE.cv(um_effect)
  um_effect_RMSE <- sqrt(user_movie_effects_MSE)
  put_log2("Function: user_movie_effects_RMSE.cv:
%1-Fold Cross Validation ultimate RMSE: %2", CVFolds_N, um_effect_RMSE)
  um_effect_RMSE
}

### Model Building
\

::: {.noteblock data-latex=""}
The full source code of builing and training the current model is available in the [Model building: User+Movie Effect](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L721) section of the [capstone-movielens.main.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R) script on _GitHub_.
:::

Below, we provide the most significant part of the code for training our model using the `5-Fold Cross Validation` method:
```{r eval=FALSE}
  cv.UM_effect <- train_user_movie_effect.cv()
```
```{r}
str(cv.UM_effect)

par(cex = 0.7)
hist(cv.UM_effect$b, 30, xlab = TeX(r'[$\hat{beta}_{j}$)]'),
     main = TeX(r'[Histogram of $\hat{beta}_{j}$]'))
```

We can now construct predictors and see how much the `RMSE` improves[@IDS2_23-5]:
```{r eval=FALSE}
cv.UM_effect.RMSE <- calc_user_movie_effect_RMSE.cv(cv.UM_effect)

RMSEs.ResultTibble.UME <- RMSEs.ResultTibble.UE |> 
  RMSEs.AddRow("User+Movie Effect Model", cv.UM_effect.RMSE)

>>>>

@@

::: {.noteblock data-latex=""}
The full source code of these functions is available in the [Logging Helper functions](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L20) section of the [capstone-movielens.main.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L20) script on _GitHub_.
:::

@@

We will split the `edx` dataset into a training set, which we will use to build and train our models, and a test set in which we will compute the accuracy of our predictions, the way described in [Section 23.1.1 Movielens data](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#movielens-data) of the _Course Textbook_ mentioned above[@IDS2_23-1-1].We will also use the _5-Fold Cross Validation_ method as described in [Section 29.6 Cross validation](https://rafalab.dfci.harvard.edu/dsbook-part-2/ml/resampling-methods.html#cross-validation) of the _Course Textbook_. 
To prepare datasets for processing, we will use the following functions, specifically designed for these operations:

::: { #func.make_source_datasets .sidebar }
  - [make_source_datasets](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R#L87)
:::

::: { #func.init_source_datasets .sidebar }
  - [init_source_datasets](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R#L209)
:::

@@

### The [make_source_datasets](#func.make_source_datasets) function
\

@@

::: {.noteblock data-latex=""}
This code snippet is a [part](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R#L140) of the [make_source_datasets](#func.make_source_datasets) _function_ code described above.
:::

Note that we used the [sample_train_validation_sets](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R#L15) function call to split the original dataset (`edx` in this case): 
```{r eval=FALSE}
    split_sets <- edx |>
      sample_train_validation_sets(fold_i*1000)
```

@@
